introduction

### 2019-03-20 introduction.Rmd
edit

Australia has changed in many ways over the last two decades. Rising house prices, country-wide improvements in education, an ageing population, and a decline in religious affiliation, are just a few facets of the country's evolving socio-demographic characteristics[^1]. At the same time, political power has moved back and forth between the two major parties. In the 2007 and 2010 federal elections, the Australian Labor Party (Labor) was victorious, whereas the 2001, 2004, 2013 and 2016 elections were won by the Liberal National coalition (Liberal). The two-party preferred vote, a measure of support between these two parties, fluctuated between 47.3% and 53.5% (in favour of the Liberal party) over this period. This study explores how electoral characteristics relate to two-party preference, and whether their effects have changed over time. Information about electoral socio-demographics are from the Census, and vote counts from the federal election.

Joining these two data sources is problematic, as there in an inherent asynchronicity in the two events - a Census occurs every five years, whereas an election usually occurs every three years. The predominant approach in previous studies is to join voting outcomes to the nearest Census, without accounting for any temporal differences (see @DavisStimson98, @Stimson06, @Liao09 and @Stimson09). Furthermore, electoral boundaries change regularly, so spatial discrepancies also arise when matching electoral data. To obtain appropriate Census data for the 2004, 2007, 2010 and 2013 elections, we adopt impute electoral socio-demographics using spatio-temporal imputation that combines areal interpolation [@Goodchild1993] and linear time-interpolation. The process of collection and wrangling is detailed in @CHAPTER2 and imputation process is outlined in @CHAPTER3. All data and documentation relating to this procedure is available in the `eechidna` `R` package on `CRAN`.

Previous work on modelling voting outcomes with aggregate socio-demographics find that relatively good models can be fit. @Forrest01 does this using multiple regression of the Liberal and Labor primary vote for polling booths in the Farrer electorate in 1998. @Stimson06, @Stimson09 and @Stimson12 conduct principal component analysis of polling booths in the 2001, 2004 and 2007 elections respectively, also finding that socio-demographic characteristics of polling booths are linked to their two-party preferred vote. @Stimson09 also model the polling booth swing vote (change in the two-party preferred vote) in the 2007 election, but had little success. Instead of analyzing a single election in isolation, this paper employs a consistent model framework across elections so that any temporal changes are observed, where each of the six federal elections are modelled with a cross-sectional data set. This consists of the two-party preferred vote, which is the response variable, and socio-demographic variables which are explanatory variables that characterise each electorate. This process is discussed in @CHAPTER4 and involves standardizing variables across years, using principal components to group variables, and then selecting the most influential variables with an information-theoretic approach via Akaike weights to determine relative variable importance [@BurnhamAnderson2002]. Spatial error models are then fit to account for spatial dependence.

@@

>>>>

Australia has changed in many ways over the last two decades. Rising house prices, country-wide improvements in education, an ageing population, and a decline in religious affiliation, are just a few facets of the country's evolving socio-demographic characteristics[^1]. At the same time, political power has moved back and forth between the two major parties. In the 2007 and 2010 federal elections, the Australian Labor Party (Labor) was victorious, whereas the 2001, 2004, 2013 and 2016 elections were won by the Liberal National coalition (Liberal). The two-party preferred vote, a measure of support between these two parties, fluctuated between 47.3% and 53.5% (in favour of the Liberal party) over this period. This study explores how electoral characteristics relate to two-party preference, and whether their effects have changed over time. Electoral socio-demographics are derived from the Census, and vote counts are obtained from the federal election.

Joining these two data sources is problematic as there in an inherent asynchronicity in the two events, a Census occurs every five years, whereas an election usually occurs every three years. The first problem addressed is that of obtaining appropriate Census data for the 2004, 2007, 2010 and 2013 elections - election years in which a Census does not occur. The predominant approach in previous studies is to join voting outcomes to the nearest Census, without accounting for any temporal differences (see @DavisStimson98, @Stimson06, @Liao09 and @Stimson09). Furthermore, electoral boundaries change regularly, so spatial discrepancies also arise when matching electoral data. To obtain appropriate Census data for these four elections, electoral socio-demographics are imputed using a spatio-temporal imputation that combines areal interpolation [@Goodchild1993] and linear time-interpolation. The process of collection and wrangling the raw data is outlined in @CHAPTER2, and the imputation process is detailed in @CHAPTER3. All data and associated documentation relating to this procedure has been made available in the `eechidna` `R` package on `CRAN`, providing a resource for future analysis.

Previous work on modelling voting outcomes with aggregate socio-demographics find that relatively good models can be fit. @Forrest01 does this using multiple regression of the Liberal and Labor primary vote for polling booths in the Farrer electorate in 1998. @Stimson06, @Stimson09 and @Stimson12 conduct principal component analysis of polling booths in the 2001, 2004 and 2007 elections respectively, also finding that socio-demographic characteristics of polling booths are linked to their two-party preferred vote. @Stimson09 model the polling booth swing vote (change in the two-party preferred vote) in the 2007 election, but had little success. Instead of analyzing a single election in isolation, this paper employs a consistent model framework across six elections so that any temporal changes are observed, where each federal elections is modelled with a cross-sectional data set. Each cross-section consists of the two-party preferred vote (response variable), and socio-demographic variables (explanatory variables) that characterise each electorate. This process involves standardizing variables, using principal components to group variables into "factors" and finally fitting a spatial error model. These steps, and the rational behind a spatial model, are discussed in @CHAPTER4. Inference is then conducted on the models and insights are discussed in @CHAPTER5.

@@

, and then selecting the most influential variables with an information-theoretic approach via Akaike weights to determine relative variable importance [@BurnhamAnderson2002]. Spatial error models are then fit to account for spatial dependence.
